Comparing vector-based and ACT-R memory models using large-scale datasets:\protect\\User-customized hashtag and tag prediction on Twitter and StackOverflow

### 2014-11-25 writeups/dissertation/dissertation.tex
foo

The core research question motivating this proposal is:
What kinds of cognitively-plausible models can best predict the tags that a user on a social media site uses?
This question can be tested by generating a prediction for the most likely used hashtag whenever a user is about to generate a hashtag, and then comparing the user's chosen hashtag to the model's prediction.
This process can be framed as a memory retrieval problem.
The process of suggesting a relevant hashtag to a user can be thought of as a memory retrieval request for a hashtag, given prior hashtag use and current context.
That is, each user on a social media site has created content that contains a set of tags within the context of each post.

Co-occurrence-based modeling has been shown as a potentially useful approach when predicting Twitter hashtag use \parencite{Efron2010}.
Several memory retrieval models are based on this broad methodology, where a count is maintained of the times each contextual element (such as a word in a post) co-occur with a tag.
At least two of the current memory retrieval models are based on a co-occurrence methodology:
ACT-R's declarative memory (DM) retrieval theory and random permutation vector-based memory systems.

@@

In order to represent word order information, a function must be used that converts an environment vector for a set of words and positions (e.g., ``stack $\Phi$'') into a new uncorrelated environment vector.
The BEAGLE model uses circular convolution and circular correlation as the encoding and decoding operations.
However, that is certainly not the only operation that can generate new uncorrelated vectors from a set of memory vectors and positions.
\textcite{Sahlgren2008} used a much simpler random permutation method for this operation.

@@

Both random permutations and LSA calculate word similarity on a reduced matrix.
For LSA, the original word by document matrix is compressed to a factor by document matrix and the number of factors is much less than the number of words.
For the StackOverflow and Twitter datasets, the number of documents are the number of different tags.

@@

Information is lost with this compression in both cases: by starting with a compressed representation with random permutations or by reducing to a compressed representation with LSA.
However, the way that the compression works in these two cases is not the same.
LSA uses SVD to find words that have highly correlated tag occurrence distributions (i.e., latent concepts), and then pools those co-occurrence counts into a single dimension.
A vector-based approach like BEAGLE or random permutations essentially randomizes the mapping of the words on the reduced set of dimensions, and does not take into account word similarity when generating the mapping.
Nonetheless, random permutations have been shown to perform similar to if not better than LSA when trained on the TASA and tested on the TOEFL \parencites{Sahlgren2008,Jones2007}.

@@

The activation of each tag ($A_{i}$) is a function of four components:
That tag's base-level activation ($B_{i}$), contextual activation for the words in the title, activation for words in the body, and an offset term ($O$).
This model has been modified from the standard ACT-R model (Table \ref{tabACTRModel}) in three ways:
Two separate contextual components were used (words in title and body of post), an offset term was added, and an entropy measure was used to compute the attentional weight ($W_{j}$).

@@

The scaled entropy measure $E_{j}$ was chosen for the attentional weight ($W_{j}$) term because it has a strong theoretical foundation and it predicts well \parencite{Dumais1991}.
It is based on Shannon's Information Theory, which measures the amount of information content (predictiveness) of a word.
So the scaled entropy ($E_{j}$) of a context word can be thought of as the amount of predictiveness that that word has to \emph{any} tag.
This means that a stop word like ``the'' will have a scaled entropy measure close to 0, while a highly-predictive word like \emph{macro} will have a scaled entropy measure close to 1.
Looking at $W_{j}$, if $E_{j}$ is low then $W_{j}$ is low, meaning that the scaled entropy measure helps allocate limited attentional resources to predictive contextual elements.

@@

An additional popular-hashtags dataset was collected for Twitter.
The purpose of this dataset is to test the accuracy of each model's predicted most likely hashtag(s) for a tweet.
The dataset contains approximately three million tweets where at least one of the 400 most popular hashtags are used.
A quota of three million tweets was selected because previously research testing Bayesian co-occurrence models on large-scale datasets has used on the order of one million documents
\parencites{Stanley2013,Douglass2010,Budiu2007,Kuo2011}.
Tweets were collected in real time until the tweet quota was reached.

@@

A custom tokenization algorithm created by \textcite{Owoputi2013} was used to chunk the words in the text.
This algorithm was specifically created for tokenizing Twitter content.
However, during testing it was also found to be robust and well suited for StackOverflow content.
The initial model results with the new tokenizer were qualitatively similar to the model results found in \textcite{Stanley2013} where the Python Natural Language Processing (NLP) toolkit \parencite{Bird2009} was used.
This is most likely because the language for the samples taken for both sites is informal English,
and it seams reasonable to think that tokenizing all of the informal language and characters included in Twitter is a much more difficult task than tokenizing most of the words used on StackOverflow.
So the same tokenization algorithm was used for both StackOverflow and Twitter, since performance was similar to \textcite{Stanley2013} for StackOverflow,
and it seemed beneficial to use the same algorithm for both datasets to make comparisons across datasets more equitable.

@@

The method of handling stop words has been shown to influence model accuracy \parencite{Sahlgren2008,Stanley2013}, so this was the first architectural concern that was examined.
\textcite{Sahlgren2008} looked at several ways to handle the commonly-occurring stop words for the random permutation model:
A data-driven frequency-weighted approach was tried, along with using a previously-generated stoplist, and no stop word removal at all.
Removing the high-frequency words in the dataset showed the highest performance increase compared to no stop word removal, and using a previously-generated stoplist hardly increased performance.
However, \citeauthor{Sahlgren2008} did not try to weight the stop words and only tried methods to remove them.
So alongside looking at frequency weighting, previously generated, and no stop word removal, \textcite{Stanley2013}'s method of attenuating stop words based on entropy was also explored.

@@

Plots for the three methods of handling stop words for the random permutation model on the StackOverflow dataset are shown in Figure \ref{figContextStopRPSO}.
The results show that model accuracy improves if stop words are handled in any of the three ways (predefined stoplist filtering, entropy weighting, and frequency filtering).
Using a predefined stoplist makes the smallest improvement (\numNoZero{.45} to \numNoZero{.49}), while the entropy (\numNoZero{.60}) and frequency (\numNoZero{.63}) techniques improve performance the most,
and frequency shows a slight edge over entropy.
However, using the frequency-weighting technique produces more noise in accuracy measurements compared to the other techniques, at least when tested on words in the title of StackOverflow posts.

@@

Model accuracy results after incorporating word order into the random permutation model are depicted in Figure \ref{figContextOrder}.
All bar plots that include an order component where ``window'' or ``direction'' is not included in the title use the standard ordering method (no direction or windowing, all position information is included) 
This is the default ordering method used, so the names for these bar plots were kept consistent to the names used in other figures so that comparisons could be made across figures if necessary
(similar to what was done for the default row size for the random permutation model).
Base models that do not use entropy weighting were also included so that effect sizes could be compared between entropy weighting and adding word order information.

>>>>

The core research question motivating this proposal is:
What kinds of cognitively-plausible models can best predict the tags that a user on a social media site uses?
This question can be tested by generating a prediction for the most likely used tag whenever a user is about to generate a tag, and then comparing the user's chosen tag to the model's prediction.
This process can be framed as a memory retrieval problem.
The process of suggesting a relevant tag to a user can be thought of as a memory retrieval request for a tag, given prior tag use and current context.
That is, each user on a social media site has created content that contains a set of tags within the context of each post.

Co-occurrence-based modeling has been shown as a potentially useful approach when predicting Twitter hashtag use \parencite{Efron2010} and StackOverflow tag use \parencites{Stanley2013}
Several memory retrieval models are based on this broad methodology, where a count is maintained of the times each contextual element (such as a word in a post) co-occur with a tag.
At least two of the current memory retrieval models are based on a co-occurrence methodology:
ACT-R's declarative memory (DM) retrieval theory and random permutation vector-based memory systems.

@@

In order to represent word order information, a function must be used that converts an environment vector for a set of words and positions (e.g., ``stack $\Phi$'') into a new uncorrelated environment vector.
The BEAGLE model uses circular convolution and circular correlation as the encoding and decoding operations.
However, that is certainly not the only operation that can generate new uncorrelated vectors from a set of memory vectors and positions.
\textcite{Sahlgren2008} used a simple random permutation method for this operation.

@@

Both vector-based models and LSA calculate word similarity on a reduced matrix.
For LSA, the original word by document matrix is compressed to a factor by document matrix and the number of factors is much less than the number of words.
For the StackOverflow and Twitter datasets, the number of documents are the number of different tags.

@@

Information is lost with this compression in both cases: by starting with a compressed representation with vector-based models or by reducing to a compressed representation with LSA.
However, the way that the compression works in these two cases is not the same.
LSA uses SVD to find words that have highly correlated tag occurrence distributions (i.e., latent concepts), and then pools those co-occurrence counts into a single dimension.
A vector-based approach like BEAGLE or random permutations essentially randomizes the mapping of the words on the reduced set of dimensions, and does not take into account word similarity when generating the mapping.
Nonetheless, random permutations have been shown to perform similar to if not better than LSA when trained on the TASA and tested on the TOEFL \parencites{Sahlgren2008,Jones2007}.

@@

The activation of each tag ($A_{i}$) is a function of four components:
That tag's base-level activation ($B_{i}$), contextual activation for the words in the title, activation for words in the body, and an offset term ($O$).
This model has been modified from the standard ACT-R model (Table \ref{tabACTRModel}) in four ways:
Two separate contextual components were used (words in title and body of post), an offset term was added, an entropy measure was used to compute the attentional weight ($W_{j}$),
and the base-level activation of a tag was based on global past tag use and not customized to a user's past tagging history.

@@

The scaled entropy measure $E_{j}$ was chosen for the attentional weight ($W_{j}$) term because it has a strong theoretical foundation \parencite{Dumais1991} and is parameter free.
It is based on Shannon's Information Theory, which measures the amount of information content (predictiveness) of a word.
So the scaled entropy ($E_{j}$) of a context word can be thought of as the amount of predictiveness that that word has to \emph{any} tag.
This means that a stop word like ``the'' will have a scaled entropy measure close to 0, while a highly-predictive word like \emph{macro} will have a scaled entropy measure close to 1.
Looking at $W_{j}$, if $E_{j}$ is low then $W_{j}$ is low, meaning that the scaled entropy measure helps allocate limited attentional resources to predictive contextual elements.

@@

An additional popular-hashtags dataset was collected for Twitter.
The purpose of this dataset was to test the accuracy of each model's predicted most likely hashtag(s) for a tweet.
The dataset contains approximately three million tweets where at least one of the 400 most popular hashtags are used.
A quota of three million tweets was selected because previously research testing Bayesian co-occurrence models on large-scale datasets has used on the order of one million documents
\parencites{Stanley2013,Douglass2010,Budiu2007,Kuo2011}.
Tweets were collected in real time until the tweet quota was reached.

@@

A custom tokenization algorithm created by \textcite{Owoputi2013} was used to chunk the words in the text.
This algorithm was specifically created for tokenizing Twitter content.
However, during testing it was also found to be robust and well suited for StackOverflow content.
The initial model results with the new tokenizer were qualitatively similar to the model results found in \textcite{Stanley2013} where the Python Natural Language Processing (NLP) toolkit \parencite{Bird2009} was used.
This is most likely because the language for the samples taken for both sites is informal English,
and it seams reasonable to think that tokenizing all of the informal language and characters included in Twitter is a much more difficult task than tokenizing most of the words used on StackOverflow.
So the same tokenization algorithm from \textcite{Owoputi2013} was used for both StackOverflow and Twitter, since performance was similar to \textcite{Stanley2013} for StackOverflow,
and it seemed beneficial to use the same algorithm for both datasets to make comparisons across datasets more equitable.

@@

The method of handling stop words has been shown to influence model accuracy \parencite{Sahlgren2008,Stanley2013}, so this was the first architectural concern that was examined.
\textcite{Sahlgren2008} looked at several ways to handle the commonly-occurring stop words for the random permutation model:
A data-driven frequency-filtering approach was tried, along with using a previously-generated stoplist, and no stop word removal at all.
Removing the high-frequency words in the dataset showed the highest performance increase compared to no stop word removal, and using a previously-generated stoplist hardly increased performance.
However, \citeauthor{Sahlgren2008} did not try to weight the stop words and only tried methods to remove them.
So alongside looking at frequency filtering, previously generated, and no stop word removal, \textcite{Stanley2013}'s method of attenuating stop words based on entropy was also explored.

@@

Plots for the three methods of handling stop words for the random permutation model on the StackOverflow dataset are shown in Figure \ref{figContextStopRPSO}.
The results show that model accuracy improves if stop words are handled in any of the three ways (predefined stoplist filtering, entropy weighting, and frequency filtering).
Using a predefined stoplist makes the smallest improvement (\numNoZero{.45} to \numNoZero{.49}), while the entropy (\numNoZero{.60}) and frequency (\numNoZero{.63}) techniques improve performance the most,
and frequency shows a slight edge over entropy.
However, using the frequency-filtering technique produces more noise in accuracy measurements compared to the other techniques, at least when tested on words in the title of StackOverflow posts.

@@

Model accuracy results after incorporating word order into the random permutation model are depicted in Figure \ref{figContextOrder}.
All bar plots that include an order component where ``window'' or ``direction'' is not included in the title use the standard ordering method (no direction or windowing, all position information is included).
This is the default ordering method used, so the names for these bar plots were kept consistent to the names used in other figures so that comparisons could be made across figures if necessary
(similar to what was done for the default row size for the random permutation model).
Base models that do not use entropy weighting were also included so that effect sizes could be compared between entropy weighting and adding word order information.
